





```{r}
names(knitr::knit_engines$get())
names(knitr::knit_engines$get())

knitr::knit_engines$get("python")

str(knitr::knit_engines$get())

knitr::knit_engines$get("bash")
```


```{r}
str(knitr::knit_engines)

knitr::knit_engines$set()
```

```{r}
knitr::opts_chunk$set(engine.path = list(
  python = '~/anaconda/bin/python',
  ruby = '/usr/local/bin/ruby'
))
```

```{bash, engine.path = "C:\\Program Files\\Git\\bin"}
ls --help
```


```{r}
# should exist
Sys.which('bash')
Sys.which('sh')
Sys.which('python')
```

```{r}
shell('setx PATH "C:\\Program Files\\Git\\bin"')
```


```{r}
Sys.which('bash')
```

```{r}
Sys.getenv("PATH")

old_path <- Sys.getenv("PATH")
old_path

C:\\Users\\ozanj\\AppData\\Local\\Programs\\Python\\Python38
#Sys.setenv(PATH = paste(old_path, "C:\\Program Files\\Git\\bin", sep = ":"))
Sys.getenv("PATH")
```

```{r test-bash, engine='bash'}
echo hello world
echo 'a b c' | sed 's/ /\|/g'
# number of lines
awk 'END{print NR;}' 027-engine-bash.Rmd
```

SHOW AND CHANGE "HOME" DIRECTORY FOR R CODE CHUNKS
```{r}
Sys.getenv("HOME")
Sys.setenv(HOME = "C:\\Users\\ozanj")
Sys.getenv("HOME")
```


We cun run `Bash` code instead of `R` code by replacing `{r}` with `{bash}` at top of chunk.

- Note that the default working directory when you run a code chunk in a .Rmd file is the directory where the .Rmd file is saved
    - This is true for an `R` code chunk, a `Bash` code chunk, or any other programming language
    - If you change working directories within a code chunk, the working directory reverts back to where the .Rmd file is saved after the code chunk finishes running
```{bash}
pwd
ls
```
