---
title: "Introduction to Git and GitHub"
author: 
date: 
urlcolor: blue
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true # toc_float option to float the table of contents to the left of the main document content. floating table of contents will always be visible even when the document is scrolled
      #collapsed: false # collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level (e.g., H2) headers. If collapsed initially, the TOC is automatically expanded inline when necessary
      #smooth_scroll: true # smooth_scroll (defaults to TRUE) controls whether page scrolls are animated when TOC items are navigated to via mouse clicks
    number_sections: true
    fig_caption: true # ? this option doesn't seem to be working for figure inserted below outside of r code chunk    
    highlight: tango # Supported styles include "default", "tango", "pygments", "kate", "monochrome", "espresso", "zenburn", and "haddock" (specify null to prevent syntax    
    theme: default # theme specifies the Bootstrap theme to use for the page. Valid themes include default, cerulean, journal, flatly, readable, spacelab, united, cosmo, lumen, paper, sandstone, simplex, and yeti.
    df_print: tibble #options: default, tibble, paged
    keep_md: true # may be helpful for storing on github
    
---

```{bash}
cd ~ # change directories to home directory

rm -rf repo_lec3  # force remove `my_git_repo` (if it exists)
mkdir repo_lec3  # make directory `my_git_repo`

# Initialize a new git repository in `my_git_repo` directory
cd repo_lec3
git init

ls -al
git status
```

add remote rep
```{bash}
cd ~/repo_lec3

ls -al

#add a remote repository that connects to local repository, thus enabling us to  push/pull 

git remote add origin https://github.com/ozanj/repo_lec3.git # https 
#git remote add ozan_origin https://github.com/ozanj/repo_lec3.git # https
#git remote add origin git@github.com:ozanj/repo_lec3.git # ssh
```

use `get remote set-url` to change url of remote
```{bash}
cd ~/repo_lec3

ls -al

#set url
#git remote set-url origin git@github.com:ozanj/repo_lec3.git

```

```{bash}
cd ~/repo_lec3

git remote

git remote -v
```

```{bash}
cd ~/repo_lec3

echo "library(tidyverse)" > lec3_script.R

git add lec3_script.R

git commit -m "initial commit, lec3_script.R"
```

note: we have committed file to local repo. we have connected local and remote repo using git remote add, but when we use git status, we don't see a note indicating that our local repo is ahead of our remote repo
```{bash}
cd ~/repo_lec3

git status
```

try pushing changes
```{bash}
cd ~/repo_lec3

git push
```

DEFINE UPSTREAM BRANCH [CONCEPTUALLY IN LECTURE]

EXPLAIN git push COMMAND [FOCUSING ON GIT PUSH BY ITSELF AND GIT PUSH --SET-UPSTREAM]

AND SHOW STUDENTS HOW THEY CAN OBTAIN THE INFO FOR:  --set-upstream <remote_name> <branch_name>

- find <remote_name>

```{bash}
cd ~/repo_lec3
git remote -v
```

- find <branch_name>

```{bash}
cd ~/repo_lec3

git branch

echo ""
#OR
git status
```

```{bash}
cd ~/repo_lec3
git push --set-upstream origin master
```




# BRANCHES

## conceptual stuff

- what is a branch

- show image of a branch

- introduce git branch command


Listing branches
```{bash}
cd ~/repo_lec3

git status

echo ""
git branch

```

Listing both local and remote branches

- Note: the branch “master” on your local machine is a different branch from “origin/master”
```{bash}
cd ~/repo_lec3

git branch -av

```


## creating branches

before we create branch, let's make a second commit on the master
```{bash}
cd ~/repo_lec3

echo "mpg %>% head(5)" >> lec3_script.R

git add lec3_script.R
git commit -m "add line to lec3_script.R"
git push
```
```{bash}
cd ~/repo_lec3
git log
```

show the commit object associated w/ commit hashes
```{bash}
cd ~/repo_lec3
echo "initial commit"
git cat-file -p fe91b419c71935c8edd3f4dcb3b6971634d681c5

echo ""
echo "2nd commit"
git cat-file -p 67bd01f2a55e4411ceeb11629d79352598782199

```

create new branch
```{bash}
cd ~/repo_lec3

git branch dev

git branch # show branches
```

delete branch
```{bash}
cd ~/repo_lec3

git branch -d dev

git branch
```

## switching to different branch [using git checkout]

introduce git checkout command

- to switch from one branch to another (existing) branch
- to create new branch and switch from existing branch to this new branch
- [add note that we also use git checkout for different purpose of discarding changes from working directory]


create branch and switch to it using git checkout
```{bash}
cd ~/repo_lec3

git branch
git checkout -b dev
git branch
```

```{bash}
cd ~/repo_lec3

git status
```

on `dev` branch, make changes to file
```{bash}
cd ~/repo_lec3

echo "mpg %>% filter(hwy >=30)" >> lec3_script.R
echo "mpg %>% filter(hwy <30)" >> lec3_script.R

git add lec3_script.R
git commit -m "add lines to lec3_script.R"

git status
```

```{bash}
cd ~/repo_lec3
git log
git cat-file -p 105cecccf921d6eb45bf26314ecaba8bd05ad732
```

## merge master and dev [no conflicts]

- define merge
- introduce terminology
“Current branch”
Branch you are currently working with
The branch will be updated/modified by the merge with the “target branch”
In the figure above, the master branch is the “current branch”
“Target branch”
Branch that will be merged into the “current branch”
Target branch will be unaffected by the merge
Often, programmers delete the target branch after merging with the current branch
In the figure above, “develop” is the target branch

show example of merging dev into master

show example of merging master into dev


## merge master and dev [with conflicts]


```{r}
library(tidyverse)
mpg %>% head(5)
mpg %>% filter(hwy >=30)
mpg %>% filter(hwy <30)
```


list branches
```{bash}
cd ~/repo_lec3

git checkout dev
git branch
```

show `git status` after you swich branches
```{r}

```

