---
title: "rvest and regex"
author: 
date: 
urlcolor: blue
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true # toc_float option to float the table of contents to the left of the main document content. floating table of contents will always be visible even when the document is scrolled
      #collapsed: false # collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level (e.g., H2) headers. If collapsed initially, the TOC is automatically expanded inline when necessary
      #smooth_scroll: true # smooth_scroll (defaults to TRUE) controls whether page scrolls are animated when TOC items are navigated to via mouse clicks
    number_sections: true
    fig_caption: true # ? this option doesn't seem to be working for figure inserted below outside of r code chunk    
    highlight: tango # Supported styles include "default", "tango", "pygments", "kate", "monochrome", "espresso", "zenburn", and "haddock" (specify null to prevent syntax    
    theme: default # theme specifies the Bootstrap theme to use for the page. Valid themes include default, cerulean, journal, flatly, readable, spacelab, united, cosmo, lumen, paper, sandstone, simplex, and yeti.
    df_print: tibble #options: default, tibble, paged
    keep_md: true # may be helpful for storing on github
    
---

# `rvest` package

> Wrappers around the `xml2` and `httr` packages to make it easy to download, then manipulate, HTML and XML.

*Source: `rvest` package documentation*

```{r, message=FALSE}
library(tidyverse)
library(rvest)
```

Why use the `rvest` package?

- `rvest` makes it easy to parse HTML
- First, we use the `read_html()` function to [read in the HTML](#reading-html) and convert it to an `xml_document`/`xml_node` object
- A **node** is just an HTML **element**
- HTML is made up of nested elements, so once we've read in the HTML to a `xml_node` object, we can easily traverse the nested nodes (ie. children elements) and [parse the HTML](#parsing-html)
- `rvest` comes with many helpful functions to search and extract various parts of the HTML
  - `html_node()`/`html_nodes()`: Search and extract node(s) (ie. HTML elements)
  - `html_text()`: Extract the content between HTML tags
  - `html_attr()`/`html_attrs()`: Extract the attribute(s) of HTML tags


## Reading HTML

__The `read_html()` function__:

```{r, eval = FALSE}
?read_html

# SYNTAX AND DEFAULT VALUES
read_html(x, encoding = "", ..., options = c("RECOVER", "NOERROR", "NOBLANKS"))
```

- Arguments:
  - `x`: The input can be a string containing HTML or url to the webpage you want to scrape
- Output:
  - The HTML that is read in will be returned as an `rvest` `xml_document`/`xml_node` object and can be easily parsed
  - You can also view the raw HTML using `as.character()`
  
__Scraping HTML from a webpage__:

- Navigate to the webpage (e.g., https://corona.help/) in your browser
  - If possible, use Google Chrome or Mozilla Firefox
- View the HTML of the page by right clicking > `View Page Source`
  - This will be the raw HTML that is scraped when we use `read_html()`
- When you right click, you may notice another option called `Inspect` (Chrome) or `Inspect Element` (Firefox) that will pop up a side panel
  - This can be helpful for visualizing the HTML elements on the page
  - You can also click on this side panel and hit `ctrl` + `f` (Windows) or `cmd` + `f` (Macs) to search for elements using a selector
  - But note that the HTML you see here might not be the same as what you see in `View Page Source` (i.e., _what is scraped_), since it also reflects changes made to the HTML _after_ the page was loaded (e.g., _by JavaScript_)

<br>
<details><summary>**Example**: Using `read_html()` to read in HTML from string</summary>

```{r}
html <- read_html("<h1>This is a heading.</h1><p>This is a paragraph.</p>")

# View object
html

# View class of object
class(html)

# View raw HTML
as.character(html)
```

</details>

<br>
<details><summary>**Example**: Using `read_html()` to scrape the page `https://corona.help/`</summary>


```{r}
corona <- read_html("https://corona.help/")

# View object
corona

# View class of object
class(corona)
```

```{r, eval=FALSE}
# View raw HTML [output omitted]
as.character(corona)
```

```{r}
# Inspect raw HTML
str(as.character(corona))
```

</details>

## Parsing HTML

__The `html_node()` and `html_nodes()` functions__:

```{r, eval = FALSE}
?html_node

# SYNTAX
html_node(x, css, xpath)


?html_nodes

# SYNTAX
html_nodes(x, css, xpath)
```

- Arguments:
  - `x`: An `rvest` `xml_document`/`xml_node` object (use `read_html()` to get this)
  - `css`: Selector (can select by HTML tag name, its attributes, etc.)
- Output:
  - `html_node()` returns the first element that it finds as an `rvest` `xml_node` object
    - Recall that a node is just an HTML element
  - `html_nodes()` returns all elements that it finds as an `rvest` `xml_nodeset` object
    - All elements that are selected will be returned in a nodeset
  - Again, you can view the raw HTML using `as.character()`
    - Syntax: `as.character(html_node(...))`

__Selecting for HTML elements__:

- HTML elements can be selected in many ways
  - Selecting by tagname: `'p'`, `'table'`, etc.
  - Selecting by class using `.`: `'.my-class'`
  - Selecting by id using `#`: `'#my-id'`
  - Selecting nested elements: `'table tr'` (_selects all rows within a table_)
- You can test your selector in your browser
  - Right click and select `Inspect` (Chrome) or `Inspect Element` (Firefox) to bring up a side panel
  - Hit `ctrl` + `f` (Windows) or `cmd` + `f` (Macs) and enter your selector to search for elements


<br>
<details><summary>**Example**: Using `html_node()` and `html_nodes()` I</summary>

Remember that the input to `html_node()`/`html_nodes()` should be an `rvest` `xml_document`/`xml_node` object, which we can obtain from `read_html()`:

```{r}
html <- read_html("<p>Paragraph #1</p><p>Paragraph #2</p><p>Paragraph #3</p>")

# View class of object
class(html)

# View raw HTML to see what elements are there
as.character(html)
```

<br>
If we search for the `<p>` element using `html_node()`, it will return the first result:

```{r}
first_p <- html_node(html, 'p')

# View class of object
class(first_p)

# View raw HTML
as.character(first_p)
```

<br>
If we search for the `<p>` element using `html_nodes()`, it will return all results:

```{r}
all_p <- html_nodes(html, 'p')

# View class of object
class(all_p)

# View raw HTML
as.character(all_p)
```

<br>
Note that we could also use `%>%`:

```{r}
# These are equivalent to the above
first_p <- html %>% html_node('p')
all_p <- html %>% html_nodes('p')
```

</details>


<br>
<details><summary>**Example**: Using `html_node()` and `html_nodes()` II</summary>


Let's revisit the HTML we scraped from https://corona.help/ in the previous example

  - We will try selecting for the "Total by country" table off of that page
  - In your browser, right click > `View Page Source` to check that the `table` element is indeed in the scraped HTML
  - Then, you can right click the table on the page and inspect it to better visualize the elements


```{r}
# Scraped HTML is stored in this `xml_document`/`xml_node` object
class(corona)
```

<br>
Select for the `<table>` element on that page using `html_node()`:

```{r}
# Since this table is the only table on the page, we can just use `html_node()`
corona_table <- corona %>% html_node('table')
corona_table

# View class of object
class(corona_table)
```

```{r, eval=FALSE}
# View raw HTML of `corona_table` [output omitted]
as.character(corona_table)
```

<br>
Select all rows in the table (i.e., `<tr>` elements) using `html_nodes()`

- It makes sense to select by row (rather than column) because each row usually represent an observation
- The way HTML tables are structured also makes it easier to extract information by row because each `<tr>` element (i.e., row) has `<th>`/`<td>` elements (i.e., column cells) nested within it, and not the other way around
- But if you wanted to select a certain column, there are ways to do that as well (e.g., `table tr td:nth-child(1)` selects the first cell in each row a.k.a. the first column in table)


```{r}
# We can chain `html_node()`/`html_nodes()` functions
corona_rows <- corona %>% html_node('table') %>% html_nodes('tr')

# Alternatively, we can use `table tr` as the selector to select all `tr` elements within a `table`
corona_rows <- corona %>% html_nodes('table tr')

# Investigate object
head(corona_rows) # View first few rows
typeof(corona_rows)
class(corona_rows)
length(corona_rows) # Number of elements
```

</details>

<br>

### Practicing regex

The following examples use the Coronavirus data from https://corona.help/

  - Recall that we have selected for all rows in the data table on that page in the previous example
  - If we wanted to try and create a dataframe out of this table, we could further select each cell in the table (i.e., `<td>` elements from each row), but this would involve loops which will be covered in a later lecture
  - For now, we will be practicing parsing data from each row using regex

View `corona_rows` we selected from previous example:

```{r}
# View first few rows
head(corona_rows)
corona_rows[1:5] # first five rows
corona_rows[c(1)] # header row
corona_rows[1] # header row
```

<br>
Let's convert this to raw HTML using `as.character()` to practice writing regular expressions. Refer back to this output to help you determine what pattern you want to match:

```{r}
# Convert rows to raw HTML
rows <- as.character(corona_rows)[-c(1)] # [-c(1)] means skip header row

# View first few rows as raw HTML
writeLines(head(rows, 2))  # printing via writeLines() is much prettier than printing via print()

# Investgate object named `rows`, which is a character vector
typeof(rows)
class(rows)
length(rows)
```

<br>
<details><summary>**Example**: Using `str_subset()` to subset rows</summary>

Let's filter for rows whose country name starts with `'United'`. First, preview what our regular expression matches using `str_view()`:

```{r}
str_view_all(string = head(rows), pattern = 'United \\w+')
```


Inspect the output from `str_detect()`, which returns `TRUE` if there is a match and `FALSE` if not. For example, we see there is a `TRUE` for the first element (United States) and third element (United Kingdom):

```{r}
str_detect(string = head(rows), pattern = 'United \\w+')
```

Finally, subset rows by country name using `str_subset()`, which keeps elements of character vector for which `str_detect()` is `TRUE` (i.e., keeps elements where the pattern "matches"):

```{r}
subset_by_country <- str_subset(string = rows, pattern = 'United \\w+')
writeLines(subset_by_country)
```


</details>

<br>
<details><summary>**Example**: Using `str_extract()` to extract link for each row</summary>

Since all links follow the same pattern, we can use regex to extract this info:

```{r}
links <- str_extract(string = rows, pattern = 'https://corona.help/country/[-a-z]+')

# View first few links
head(links)
```

</details>

<br>
<details><summary>**Example**: Using `str_match()` to extract country for each row</summary>

Since all countries are in a `div` element with the same attributes, we can use the following regex to extract the country name:

```{r}
countries <- str_match(string = rows, pattern = '<div style="height:100%;width:100%">([\\w ]+)</div>')

# View first few countries
# We used a capturing group to extract the country name from between the tags
head(countries)
```

</details>

<br>
<details><summary>**Example**: Using `str_match_all()` to extract number deaths and critical for each row</summary>

Since both the number of deaths and critical are in a `<td>` element with the same `class` attribute, we can use the following regex to extract both numbers:

```{r}
num_danger <- str_match_all(string = rows, pattern = '<td class="text-danger">([\\d,]+)</td>')

# View matches for first few rows
# We used a capturing group to extract the numbers from between the tags
head(num_danger)
```

</details>

<br>
<details><summary>**Example**: Using `str_replace_all()` to convert numeric values to thousands for each row</summary>

Rewrite all numeric values greater than one thousand in terms of `k`:

```{r}
num_to_k <- str_replace_all(string = rows, pattern = '>([\\d,]+),\\d{3}<', replacement = '>\\1k<')

# View replacements for first few rows
writeLines(head(num_to_k))
```

</details>
